### 2018-03-29 submission/manuscript.Rmd
final re-work intro & discussion read through -- hopefully Pat likes it

**Background.** Using different reagents and kits at different steps of the 16S rRNA gene sequencing workflow can introduce bias by changing the observed microbial community. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and HiFi DNA polymerase can change the bacterial community and introduce bias to the final obtained results.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for both fecal samples and mock communities. It also assessed HiFi DNA polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence. Finally, our analysis also examined chimera prevalence correlation with the number of OTUs.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011]. Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.

@@

There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index. Using this metric our observations found no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

@@

Methods can be standardized but they commonly contain bias that is reproducible and may miss important associations. Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016]. This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases may work better in certain situations over others. Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

>>>>

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Commonly studied sources of bias include preservation and DNA extraction methods. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and HiFi DNA polymerase can change the bacterial community and introduce bias to the final obtained results.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for both fecal samples and mock communities. It also assessed HiFi DNA polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence. Our analysis also examined chimera prevalence correlation with the number of OTUs.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011]. Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.

@@

There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index. Using this metric our observations found no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

@@

Methods can be standardized but they commonly contain bias that is reproducible and may miss important associations. Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016]. This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases may work better in certain situations over others. Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility and replicability more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

### 2018-03-29 submission/manuscript.tex
final re-work intro & discussion read through -- hopefully Pat likes it

\textbf{Background.} Using different reagents and kits at different
steps of the 16S rRNA gene sequencing workflow can introduce bias by
changing the observed microbial community. Although cycle number and
high fidelity (HiFi) DNA polymerase are studied less often, they are
still important sources of bias in this workflow. Here, we examine how
both cycle number and HiFi DNA polymerase can change the bacterial
community and introduce bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for both fecal samples and mock communities. It also assessed
HiFi DNA polymerase dependent differences in the Bray-Curtis index,
error rate, sequence error prevalence, and chimera prevalence. Finally,
our analysis also examined chimera prevalence correlation with the
number of OTUs.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as cycle number increases which can drastically change commonly used
diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will also
artificially increase community diversity (Haas et al., 2011). Although
these differences are not necessarily dependent on primer and DNA
polymerase used, there are also biases that are.

@@

There were few differences that depend on HiFi DNA polymerase for
between sample diversity, as measured by the Bray-Curtis index. Using
this metric our observations found no differences in the overall
bacterial community composition for sub-sampling depth or cycle number
used. One possible reason for this outcome was that our study did not
have enough power to detect differences due to low sample number in each
group. Another reason was that many of the OTUs are likely not highly
abundant, allowing the Bray-Curtis index to be able to successfully
down-weight chimeric OTUs (Minchin, 1987). The choice of downstream
diversity metric could be an important consideration in helping to
mitigate these observed HiFi DNA polymerase dependent differences in
chimera prevalence. Metrics that solely use presence/absence of OTUs
(e.g.~Jaccard (Real \& Vargas, 1996)) may be less robust to chimera
prevalence and by extension total number of OTU differences in HiFi DNA
Polymerases. When choosing a distance metric careful consideration of
the biases introduced from the PCR step of the 16S rRNA gene sequencing
workflow need to be taken into account. With differences in the number
of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it
might be easier to avoid specific DNA polymerase families altogether.

@@

Methods can be standardized but they commonly contain bias that is
reproducible and may miss important associations. Bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
2016; Amir et al., 2017). This study shows that specific diversity
metrics used to measure the microbial community consistently vary based
on HiFi DNA polymerase. Standardizing multiple workflows to one specific
HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases
may work better in certain situations over others. Arguably, the degree
of workflow standardization across studies and research group needs to
be approached on a study by study basis and not every project needs to
use the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.

>>>>

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Commonly studied sources of bias include
preservation and DNA extraction methods. Although cycle number and high
fidelity (HiFi) DNA polymerase are studied less often, they are still
important sources of bias in this workflow. Here, we examine how both
cycle number and HiFi DNA polymerase can change the bacterial community
and introduce bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for both fecal samples and mock communities. It also assessed
HiFi DNA polymerase dependent differences in the Bray-Curtis index,
error rate, sequence error prevalence, and chimera prevalence. Our
analysis also examined chimera prevalence correlation with the number of
OTUs.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as cycle number increases which can drastically change commonly used
diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will
artificially increase community diversity (Haas et al., 2011). Although
these differences are not necessarily dependent on primer and DNA
polymerase used, there are also biases that are.

@@

There were few differences that depend on HiFi DNA polymerase for
between sample diversity, as measured by the Bray-Curtis index. Using
this metric our observations found no differences in the overall
bacterial community composition for sub-sampling depth or cycle number
used. One possible reason for this outcome was that our study did not
have enough power to detect differences due to low sample number in each
group. Another reason was that many of the OTUs are likely not highly
abundant, allowing the Bray-Curtis index to be able to successfully
down-weight chimeric OTUs (Minchin, 1987). The choice of downstream
diversity metric could be an important consideration in helping to
mitigate these observed HiFi DNA polymerase dependent differences in
chimera prevalence. Metrics that solely use presence/absence of OTUs
(e.g.~Jaccard (Real \& Vargas, 1996)) may be less robust to chimera
prevalence and by extension total number of OTU differences in HiFi DNA
Polymerases. When choosing a distance metric, careful consideration of
the biases introduced from the PCR step of the 16S rRNA gene sequencing
workflow need to be taken into account. With differences in the number
of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it
might be easier to avoid specific DNA polymerase families altogether.

@@

Methods can be standardized but they commonly contain bias that is
reproducible and may miss important associations. Bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
2016; Amir et al., 2017). This study shows that specific diversity
metrics used to measure the microbial community consistently vary based
on HiFi DNA polymerase. Standardizing multiple workflows to one specific
HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases
may work better in certain situations over others. Arguably, the degree
of workflow standardization across studies and research group needs to
be approached on a study by study basis and not every project needs to
use the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility and replicability more difficult it will help to avoid
systematic biases from occuring due to widespread standardization of
approaches.
